---
output:
  md_document:
    variant: markdown_github
---

[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/eeptools)](https://cran.r-project.org/package=eeptools)
[![Downloads](https://cranlogs.r-pkg.org/badges/eeptools)](https://cran.r-project.org/package=eeptools)


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/readme/README-"
)
```

# Introduction

`eeptools` is an R package that seeks to make it easier for analysts at 
state and local education agencies to analyze and visualize their data on student, 
school, and district performance. By putting simple wrappers around a number of 
R functions, `eeptools` strives to make many common tasks simpler and less prone 
to error specific to analysis of education data.

# Datasets

`eeptools` provides three new datasets of interest to education researchers. These 
datasets are also used in the [R Bootcamp for Education Analysts](https://www.jaredknowles.com/r-bootcamp)

```{r}
library(eeptools)
data("stuatt")
head(stuatt)
```

The `stuatt`, student attributes, dataset is provided from the [Strategic Data 
Project Toolkit for Effective Data Use](https://sdp.cepr.harvard.edu/toolkit-effective-data-use). 
This dataset is useful for learning how to clean data in R and how to aggregate 
and summarize individual unit-record data into group-level data. 

```{r}
data(stulevel)
head(stulevel)

```

The `stulevel` dataset is a simulated student-level longitudinal record. It contains 
student and school level attributes and is useful for practicing evaluating 
longitudinal analyses of student unit-record data. 

```{r}
data("midsch")
head(midsch)

```

The `midsch` dataset contains an analysis on abnormality in school average 
assessment scores. It contains observed and predicted values of aggregated 
test scores at the school level for a large midwestern state. 

# Administrative Data Functions

For analysts using unit-record data of some type, there are several `calc` functions 
which automate common tasks including calculating ages (`age_calc`), 
grade retention (`retained_calc`), and student mobility (`moves_calc`). 


```{r}
age_calc(dob = as.Date('1995-01-15'), enddate = as.Date('2003-02-16'), 
         units = "years")
age_calc(dob = as.Date('1995-01-15'), enddate = as.Date('2003-02-16'), 
         units = "months")
age_calc(dob = as.Date('1995-01-15'), enddate = as.Date('2003-02-16'), 
         units = "days")
```

`age_calc` also now properly accounts for leap years and leap seconds by default. 

`retained_calc` takes a vector of student identifiers and a vector of grades and 
checks whether or not the student was retained in the grade level specified by 
the user. It returns a data.frame of all students who could have been retained 
and a yes or no indicator of whether they were retained. 


```{r}
x <- data.frame(sid = c(101, 101, 102, 103, 103, 103, 104, 105, 105, 106, 106),
                 grade = c(9, 10, 9, 9, 9, 10, 10, 8, 9, 7, 7))
retained_calc(df = x, sid = "sid", grade = "grade", grade_val = 9)
```

`retained_calc` is intended to be used after you have processed your data as it 
does not take into account time or sequence other than the order in which the 
data is passed to it. 

`moves_calc` is intended to identify based on enrollment dates whether a student 
experienced a school move within a school year. 


```{r}
df <- data.frame(sid = c(rep(1,3), rep(2,4), 3, rep(4,2)),
                   schid = c(1, 2, 2, 2, 3, 1, 1, 1, 3, 1),
                   enroll_date = as.Date(c('2004-08-26',
                   '2004-10-01', '2005-05-01', '2004-09-01',
                   '2004-11-03', '2005-01-11', '2005-04-02',
                   '2004-09-26', '2004-09-01','2005-02-02'), format='%Y-%m-%d'),
                   exit_date = as.Date(c('2004-08-26', '2005-04-10',
                    '2005-06-15', '2004-11-02', '2005-01-10',
                    '2005-03-01', '2005-06-15', '2005-05-30',
                    NA, '2005-06-15'), format='%Y-%m-%d'))

moves <- moves_calc(df, sid = "sid", schid = "schid", enroll_date = "enroll_date", 
                    exit_date = "exit_date")
moves

```

# Manipulate Data

Another set of key functions in the package are to make basic data manipulation 
easier. One thing users of other statistical packaegs may miss when using R is 
a convenient function for determining the `mode` of a vector. The `statamode` 
function is designed to do just that. `statamode` works with numeric, character,
 and factor data types. It also includes various options for how to deal with a 
 tie demonstrated below.

```{r statamode}
vecA <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
statamode(vecA, method = "stata")
vecB <- c(1, 1, 1, 3:10)
statamode(vecB, method = "last")
vecC <- c(1, 1, 1, NA, NA, 5:10)
statamode(vecC, method = "last")
vecA <- c(LETTERS[1:10]); vecA <- factor(vecA)
statamode(vecA, method = "last")
vecB <- c("A", "A", "A", LETTERS[3:10]); vecB <- factor(vecB)
statamode(vecB, method = "last")
vecA <- c(LETTERS[1:10])
statamode(vecA, method = "sample")
vecB <- c("A", "A", "A", LETTERS[3:10])
statamode(vecB, method = "stata")
vecC <- c("A", "A", "A", NA, NA, LETTERS[5:10])
statamode(vecC, method = "stata")
```

There are a number of functions to save you keystrokes like `defac` for converting 
a factor to a character, `makenum` for turning a factor variable into a numeric 
variable, `max_mis` for taking the maximum of a vector of numerics and ignoring 
any NAs (useful for inclusion in `do.call` or `apply` constructions). `remove_char` 
allows you to quickly `gsub` out a specific character from a string vector such 
as an `*` or `...`. `decomma` is a somewhat specialized version of this for 
processing data where numerics are written with commas. `nth_max` allows you to 
identify the 2nd, 3rd, etc. maximum value in a vector. 


# Regression Models

`eeptools` includes ways to simplify the use of regression analyses tools recommended 
by Gelman and Hill 2006 through the `gelmansim` function, which itself is a wrapper 
for the `arm::sim()` function. 

```{r sim}
require(MASS)
#Examples of "sim" 
set.seed (1)
J <- 15
n <- J*(J+1)/2
group <- rep (1:J, 1:J)
mu.a <- 5
sigma.a <- 2
a <- rnorm (J, mu.a, sigma.a)
b <- -3
x <- rnorm (n, 2, 1)
sigma.y <- 6
y <- rnorm (n, a[group] + b*x, sigma.y)
u <- runif (J, 0, 3)
dat <- cbind (y, x, group)
# Linear regression 
dat <- as.data.frame(dat)
dat$group <- factor(dat$group)
M3 <- glm (y ~ x + group, data=dat)
cases <- expand.grid(x = seq(-2, 2, by=0.1), 
                     group=seq(1, 14, by=2))
cases$group <- factor(cases$group)
sim.results <- gelmansim(mod = M3, newdata = cases, n.sims=200, na.omit=TRUE)
head(sim.results)
```


There is also a `ggplot2` version of `plot.lm` included:

```{r lmautoplot}
data(mpg)
mymod <- lm(cty~displ + cyl + drv, data=mpg)
autoplot(mymod)
```

Finally, there is a convenient method for creating labeled mosaic plots. 

```{r crossplot}
sampDat <- data.frame(cbind(x=seq(1,3,by=1), y=sample(LETTERS[6:8], 60, 
                                                        replace=TRUE)),
                        fac=sample(LETTERS[1:4], 60, replace=TRUE))
varnames<-c('Quality','Grade')
crosstabplot(sampDat, "y", "fac", varnames = varnames,  label = TRUE, 
             title = "Crosstab Plot", shade = FALSE)
```

# Helping Out

Review the [Contributor Guide](https://github.com/jknowles/eeptools/blob/master/CONTRIBUTING.md) for specific directions and tips 
on how to get involved.

`eeptools` is intended to be a useful project for the education analytics 
community. Contributions are welcomed. Please note that this project is 
released with a 
[Contributor Code of Conduct](https://github.com/jknowles/eeptools/blob/master/CONDUCT.md). By participating in this project you 
agree to abide by its terms.

